---
title       : Two-Way ANOVA
subtitle    : Chapter 3
author      : Stat 217
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow      # 
widgets     : [bootstrap, quiz, mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---
```{r setup, echo=FALSE, message=FALSE}
require(knitr)
require(ggplot2)
options(show.signif.stars = FALSE)
opts_chunk$set(tidy = FALSE, message=FALSE, comment = "", echo = TRUE, 
               fig.align = 'center', fig.height = 5)
tennis <- read.csv("tennis.csv", header = T)
set.seed(2014)
```

## Tennis Ball Data

Recall that in class six groups collected information on tennis ball diameter using three different methods.  The data are summarized below.  We'll use a significance level of 0.05 for all tests on these data.

```{r comment = ""}
summary(tennis)
```


---
## Warpbreaks Data

For some reason, people were interested in the number of warp breaks per loom (I have no idea what a warp break is; that's the fun of being a statistician).  They used two different types of wool (A or B) and set the looms to one of three different tensions (Low, Medium, or High).  The counts in each wool-tension group are displayed below.  We would like to find out if any wool-tension combination results in fewer breaks than the others.

```{r warp_tally, comment = ""}
require(mosaic)
tally(~ wool + tension, data = warpbreaks)
```

--- &radio
## Information about the Warpbreaks Data

Are the warp breaks data balanced?

1. _Yes_
2. No
3. Impossible to tell

*** .hint
Looks back at the counts in each wool-tension group

*** .explanation
Each wool-tension group consists of 9 looms so the data are balanced.

---
## Boxplots of the Tennis Ball Data

Separate boxes are given for each method-group combination.  The measurements for Cool Group Name have by far the most variability.  It appears that the pencil and string measurements have slightly larger diameters than the zip-tie measurements

```{r tennis_box, echo=FALSE, fig.align='center', fig.height=6, fig.width=9}
qplot(method, diameter, data = tennis, fill = method, geom = "boxplot") + facet_wrap(~Group) + 
  scale_fill_discrete(name = "Method") + 
  labs(x = "Method", 
       y = "Diameter (cm)") + theme_bw()
```

---
## Interaction Plots of the Tennis Ball Data

An interaction plot of the tennis ball data is below.  This plot is pretty messy.  The lines are not parallel all the way across and they do cross at several places which is suggestive of an interaction.  The lines do follow a general pattern though.  Overall, I would say this plot does not provide strong evidence one way or another of an interaction.  We'll have to do an F test to be sure.

```{r tennis_int, fig.align = 'center', fig.height=5}
with(tennis, interaction.plot(Group, method, diameter))
```

---
## Interaction Plot of the Warpbreaks Data

The interaction plot for the warpbreaks data are provided below.  The different lines correspond to different wool types.

```{r warp_int, fig.align = 'center', fig.height=5}
with(warpbreaks, interaction.plot(tension, wool, breaks))
```

--- &radio

## Interaction for Warpbreaks Data?

Does the interaction plot suggest that the effect of tension depends on the type of wool?  You might want to take a look at the explanation even if you get the question right.  I think it's useful.

1. Yes - the lines in the interaction plot are parallel
2. No - the lines in the interaction plot are parallel
3. _Yes - the lines in the interaction plot are not parallel_
4. No - the lines in the interaction plot are not parallel

*** .hint
Take a look back at the interaction plot.  Is there a constant shift from wool A to wool B for all levels of tension?

*** .explanation
At low tension, wool A has a lot more breaks than wool B.  At medium tension, however, wool B has more breaks.  Wool A has more breaks once again at high tension.  Therefore, the effect of wool A depends on the tension so there is probably an interaction present.

--- 
## Interaction Model for Tennis Ball Data

We'll start by fitting a model with and interaction and testing to see if the interaction is needed.  The interaction model is:

$y_{ijk} = \alpha + \tau_j + \gamma_k + \omega_{jk} + \epsilon_{ijk}$

We'll let the $\tau_j$'s be the method effects and the $\gamma_k$'s be the group effects.

To test for an interaction, we will test to see if the $\omega_{jk}$'s are all zero.

--- &radio
## Interaction Model for Warpbreaks Data

We'll also start with an interaction model for the warpbreaks data:

$y_{ijk} = \alpha + \tau_j + \gamma_k + \omega_{jk} + \epsilon_{ijk}$

We'll let the $\tau_j$'s be the tension coefficients, the $\gamma_k$'s be the wool coefficients, and the $\omega_{jk}$'s be the interaction coefficients.

Which null hypothesis tests to see if an interaction is present in the data?

1. $H_0: \tau_1 = \tau_2 = \tau_3 = 0$
2. $H_0: \gamma_1 = \gamma_2 = 0$
3. _$H_0: \omega_{11} = \omega_{12} = \ldots = \omega_{JK} = 0$_
4. $H_0: \mu_{11} = \mu_{12} = \ldots = \mu_{JK}$

*** .hint 
Which coefficients correspond to the interaction?

*** .explanation
The $\omega_{jk}$'s are the coefficients for the interaction terms.  The alternative hypothesis is:

$H_A:$ at least one $\omega_{jk}$ is not equal to zero

---
## ANOVA Table for Tennis Ball Data Part 1

Here's a sparse ANOVA table for the tennis ball data that has not been filled in.  

|id            |  Df|     Sum Sq|    Mean Sq|   F value|     Pr(>F)|
|:-------------|---:|----------:|----------:|---------:|----------:|
|method        |$df_A$|   $SS_A$|   $MS_A = SS_A / df_A$  |$F_{A} = MS_A / MS_E$ |          |
|Group         |$df_B$|   $SS_B$|   $MS_B= SS_B / df_B$  |     $F_B = MS_B / MS_E$|          |
|method:Group  |$df_{AB}$|$SS_{AB}$|$MS_{AB}= SS_{AB} / df_{AB}$| $F_{AB} = MS_{AB} / MS_E$|          |
|Residuals     |$df_{E}$| $SS_E$|$MS_E= SS_E / df_E$     |          |          |
|Total         |$df_T$| $SS_{Total}$|       |          |           |

$df_A = J-1 = 3-1$ 

$df_B = K-1 = 6-1$

$df_{AB} = (J-1) * (K-1) = (3-1) * (2-1)$

$df_E = N - (J * K) = 81 - (6 * 3)$

$df_{Total} = N-1 = 81 - 1$

---
## ANOVA Table for the Warpbreaks Data Part 1

Here's a blank ANOVA Table for a model with an interaction for the warpbreaks data.  We'll let tension be factor A and wool be factor B

|id            |  Df|     Sum Sq|    Mean Sq|   F value|     Pr(>F)|
|:-------------|---:|----------:|----------:|---------:|----------:|
|tension       |   |  |  |  |  |
|wool          |   |   |   |  |  |
|tension:wool  |   |  |   |  |  |
|Residuals     |  |  |   |      |         |
|Total | | | | |

--- &multitext
## Degrees of Freedom for Warpbreaks Data

1. What is $df_A$?
2. What is $df_B$?
3. What is $df_{AB}$?
4. What is $df_E$?
5. What is $DF_{Total}$?

*** .hint
The formulas for the degrees of freedom are on the previous slide

*** .explanation
1. <span class="answer">2</span>
There are three levels of tension so $DF_A = 3-1$

2. <span class="answer">1</span>
There are three levels of wool so $DF_B = 2-1$

3. <span class="answer">2</span>
$DF_{AB} = (J-1) * (K-1) = (3-1) * (2-1)$

4. <span class="answer">48</span>
$DF_{E} = N - J*K  = 54- 3*2$

5. <span class="answer">53</span>
$DF_{Total} = N-1 = 54-1$


---
## ANOVA Table for Tennis Ball Data Part 2

Here's the updated ANOVA table with a few numbers added to it.  We can fill in the rest of the SS and MS values.

|id            |  Df|     Sum Sq|    Mean Sq|   F value|     Pr(>F)|
|:-------------|---:|----------:|----------:|---------:|----------:|
|method        |2   |       0.84|   $MS_A = SS_A / df_A$  |$F_{A} = MS_A / MS_E$ |          |
|Group         |5   |     $SS_B$|      0.340|     $F_B = MS_B / MS_E$|          |
|method:Group  |2   |  $SS_{AB}$|      0.076| $F_{AB} = MS_{AB} / MS_E$|          |
|Residuals     |63  |       3.83|$MS_E= SS_E / df_E$     |          |          |
|Total         |80  | $SS_{Total}$|       |          |           |

$SS_B = MS_B * df_B = 0.340*5 = 1.7$

$SS_{AB} = MS_{AB} * df_{AB} = 0.076 * 2 = 0.15$

$SS_{Total} = SS_A + SS_B + SS_{AB} + SS_E = 7.13$

$MS_A = SS_A / df_A = 0.84/2 = 0.42$

$MS_E = SS_E / df_E = 3.83 / 63 = 0.061$

---
## ANOVA Table for Warpbreaks Data Part 2

A few numbers have been added to the warpbreaks ANOVA table as well.  We'll try and fill in the rest of the table.

|id            |  Df|     Sum Sq|    Mean Sq|   F value|     Pr(>F)|
|:-------------|---:|----------:|----------:|---------:|----------:|
|tension       | 2  |  2034.3|  $MS_A$|  |  |
|wool          |1   |   450.7|   $MS_B$|  |  |
|tension:wool  |2   | $SS_{AB}$ |   501.39|  |  |
|Residuals     |48  | $SS_E$ |   119.69|      |         |
|Total |53     | $SS_{Total}$| | |

--- &multitext
## SS and MS for Warpbreaks Data

Fill in the following values.  Round all answers to 2 decimal places.

1. $SS_{AB} =$
2. $SS_E =$
3. $SS_{Total}=$
4. $MS_A =$
5. $MS_B=$

*** .hint
The formulas are on the previous slide

*** .explanation
1. <span class="answer">1002.78</span>
$SS_{AB} = MS_{AB} * df_{AB} = 501.39*2$
2. <span class="answer">5745.12</span>
$SS_{E} = MS_{E} * df_{E} = 119.69*48$
3. <span class="answer">9232.9</span>
$SS_{Total} = SS_A + SS_B + SS_{AB} + SS_E$
4. <span class="answer">1017.15</span>
$MS_{A} = SS_{A} / df_{A} = 2034.3/2$
5. <span class="answer">450.</span>
$MS_{B} = SS_{B} * df_{B} = 450.7/1$

---
## ANOVA Table for Tennis Ball Data Part 3

The F-statistics for each of the three rows are just the $MS$ for that row divided by the $MS_E$.

|id            |  Df|     Sum Sq|    Mean Sq|   F value|     Pr(>F)|
|:-------------|---:|----------:|----------:|---------:|----------:|
|method        |2   |       0.84|   0.418   |$F_{A} = MS_A / MS_E$ |          |
|Group         |5   |     1.7   |      0.340|     $F_B = MS_B / MS_E$|          |
|method:Group  |2   |      0.15 |      0.076| $F_{AB} = MS_{AB} / MS_E$|          |
|Residuals     |63  |       3.83| 0.061     |          |          |
|Total         |80  | 7.13      |       |          |           |

$F_A = 0.418 / 0.061 = 6.88$

$F_B = 0.340 / 0.061 = 5.57$

$F_{AB} = 0.076 / 0.061 = 1.25$

--- 
## ANOVA Table for Warpbreaks Data Part 3

Here is the current ANOVA table for Warpbreaks data.  Next we'll fill in the F-statistics

|id            |  Df|     Sum Sq|    Mean Sq|   F value|     Pr(>F)|
|:-------------|---:|----------:|----------:|---------:|----------:|
|tension       | 2  | 2034.3 |  1017.15|  |  |
|wool          |1   |  450.7 |  450.7 |  |  |
|tension:wool  |2   | 1002.78 | 501.39  |  |  |
|Residuals     |48  | 5745.12 | 119.69  |      |         |
|Total |53 |9232.9 | | |

--- &multitext
## F-statistics for Warpbreaks Data

Fill in the following.  Round to two decimal places.

1. $F_A =$
2. $F_B =$
3. $F_{AB}=$

*** .hint
The formulas are given two slides ago

*** .explanation
1. <span class="answer">8.50</span>
$F_A = MS_A / MS_E = 1017.15/119.69$
2. <span class="answer">3.77</span>
$F_B = MS_B / MS_E = 450.7/119.69$
3. <span class="answer">4.19</span>
$F_{AB} = MS_AB / MS_E = 501.39/119.69$

---
## ANOVA Table for Tennis Ball Data Part 4

The F-statistics still follow an F distribution.  The numerator degrees of freedom is the df on the same row as each statistic.  The denominator degrees of freedom is the Residual df ($df_E$).  We compare the observed F-values to these named distributions to find a p-value.

|id            |  Df|     Sum Sq|    Mean Sq|   F value|     Pr(>F)|
|:-------------|---:|----------:|----------:|---------:|----------:|
|method        |2   |       0.84|   0.418   |  6.88 |          |
|Group         |5   |     1.7   |      0.340|     5.59 |          |
|method:Group  |2   |      0.15 |      0.076| 1.25   |          |
|Residuals     |63  |       3.83| 0.061     |          |          |
|Total         |80  | 7.13      |       |          |           |

Formally:

$F_A \sim F(2, 63)$

$F_B \sim F(5, 63)$

$F_{AB} \sim F(2, 63)$

---
## Testing for an Interaction in Tennis Ball Data

We start by first examining the method:group row.  This row tests to see if there is an interaction present between Group and method.

The F-statistic is 1.25.  This test statistic follows an F distribution with 10 numerator and 63 denominator degrees of freedom.  This yields a p-value of 0.28. We therefore fail to reject the null and conclude that the differences in diameter measurements do not vary across the groups (i.e. we conclude that there is not an interaction effect in the data).

```{r tennis_anova1, comment="", eval=FALSE}
tennis1 <- lm(diameter ~ method + Group + method*Group, data = tennis)
anova(tennis1)
```

```{r tennis_anova1_print, comment="", echo=FALSE, results='asis'}
df1 <- data.frame(anova(tennis1))
df1[, 2] <- round(df1[, 2], 3)
df1[, 3] <- round(df1[, 3], 3)
df1[, 4] <- round(df1[, 4], 3)
df1[, 5] <- round(df1[, 5], 4)
df1[4, 4:5] <- ""
names(df1) <- c("Df", "Sum Sq", "Mean Sq", "F value", "Pr(>F)")
kable(df1, format = "markdown", digits = 3)
```

--- 
## Completed ANOVA Table for Warpbreaks Data

Here is the completed ANOVA table for the Warpbreaks data. 

```{r warp_anova, comment=""}
warp.lm <- lm(breaks ~ tension + wool + tension*wool, data = warpbreaks)
anova(warp.lm)
```

--- &radio
## Beginning to Test Warpbreaks Data

Which null hypotheses is being tested by the `wool:tension` row of the ANOVA table?  Make sure that you know what the alternative hypothesis is.

1. $H_0: \tau_1 = \tau_2 = \ldots = \tau_J = 0$
2. $H_0: \gamma_1 = \gamma_2 = \ldots = \gamma_K = 0$
3. _$H_0: \omega_{11} = \omega_{12} = \ldots = \omega_{JK} = 0$_
4. $H_0: \mu_{11} = \mu_{12} = \ldots = \mu_{JK}$

*** .hint
Which greek coefficient corresponds to an interaction?

*** .explanation
The $\omega$'s are used to represent the interaction effect.  So we want to test to see if they are equal to zero.  The alternative hypothesis is 

$H_A:$ at least one $\omega_{jk} \neq =$

--- &radio
## Testing for an Interaction in Warpbreaks data

Based on the ANOVA table, is there evidence of an interaction between wool and tension at the $\alpha = 0.05$ significance level?

1. Yes
2. No

*** .hint
Check the p-value testing whether there is an interaction

*** .explanation
The p-value testing for an interaction is 0.021 so we reject the null of no interaction and conclude that there is an interaction present.

---
## Next step for Tennis Ball Data
Since we did not find any evidence of an interaction, the next step is to fit an additive model.  In this model, we say that $y_{ijk} = \alpha + \tau_j + \gamma_k + \epsilon_{ijk}$.  The ANOVA table for this model is displayed below.  This table uses Type II Sums of Squares so we can test both factors simultaneously.

```{r tennis_anova2, comment="", message=FALSE}
require(car) # for Type II sums of squares
tennis2 <- lm(diameter ~ method + Group, data = tennis)
Anova(tennis2)
```

--- &radio
## Hypotheses for Additive Model

We'll use the additive model model where the $\tau_j$'s are the method (string/ziptie/pencil) coefficients and the $\gamma_k$'s are the coefficients for different groups.  The method row tests whether the $\tau_j$'s are equal to zero (i.e. whether the different methods result in different diameters).

What null hypothesis is being tested by the p-value in the `Group` row?

1. $H_0: \tau_1 = \tau_2 = \ldots = \tau_J = 0$
2. _$H_0: \gamma_1 = \gamma_2 = \ldots = \gamma_K = 0$_
3. $H_0: \omega_{11} = \omega_{12} = \ldots = \omega_{JK} = 0$
4. $H_0: \mu_{11} = \mu_{12} = \ldots = \mu_{JK}$

*** .hint 
Which coefficients correspond to the different groups?

*** .explanation
The $\gamma_k$'s are the coefficients for the different methods so the `method` row tests to see if they are zero.

---
## Checking Equal Variance for Tennis Ball Data

To check equal variance, we look at residual vs fitted value and scale-location plots as before.  Once again, we are looking for any sort of pattern.  You could claim there's evidence of a fan shape in the residuals vs fitted values plot, but that's mostly driven by point 70 which is an outlier (we'll learn how to deal with them later).  Otherwise, the assumption appears to be okay.  These plots are strange so ask me if you are confused.

```{r tennis_equ, fig.align='center', fig.height=4, fig.width=8}
par(mfrow=c(1,2))
plot(tennis2, which = c(1,3))
```

---
## Checking Equal Variance for the Warpbreaks Data

Here are the Residuals vs Fitted values and Scale-Location plots for the warpbreaks data.

```{r warp_equ, fig.align='center', fig.height=4, fig.width=8}
par(mfrow=c(1,2))
plot(warp.lm, which = c(1,3))
```

--- &radio
## Comparing Equal Variance for the Two Data sets

For which data set does the equal variance assumption appear to be more problematic?

1. _Warpbreaks_
2. Tennis ball
3. Neither data set has a problem with the equal variance assumption
4. The equal variance assumption is violated for both data sets

*** .hint
Do you notice any patterns in the warpbreaks plots?

*** .explanation
There is an increasing trend in the Scale-Location plot for the warpbreaks data.  The residuals for the small fitted values in the warpbreaks data are all between -10 and 10; for the larger fitted values, however, the residuals range from -20 to 30.  So there is probably non-constant variance in the warpbreaks data.

---
## Checking Normality for Tennis ball Data

We can use Normal Q-Q plots and a histogram to assess normality once again.  Except for observations 70 and 71, the points fall really close to the dashed lines.  This suggests that these points are potential outliers.  The histogram looks unimodal and reasonably symmetric.  So these plots suggest that the normality assumption is met.

```{r tennis_norm, fig.align='center', fig.height=3.5, fig.width=7}
par(mfrow = c(1,2))
plot(tennis2, which = 2)
eij <- resid(tennis2)
hist(eij, main = "Histogram of Residuals")
```

---
## Checking the Normality Assumption for the Warpbreaks Data

Here are the Residuals vs Fitted values and Scale-Location plots for the warpbreaks data.

```{r warp_nrom, fig.align='center', fig.height=4, fig.width=8}
par(mfrow=c(1,2))
plot(warp.lm, which = c(2))
eij <- resid(warp.lm)
hist(eij, main = "Histogram of Residuals")
```

--- &radio
## Comparing Normality for the Two data Sets

Which data set seems to have more of a problem with the normality assumption?

1. Warpbreaks
2. Tennis ball
3. The normality assumption is not met for either data set
4. _The normality assumption is okay for both data sets_

*** .hint
Recall that we have a fairly large sample size and a balanced design for the warpbreaks data.

*** .explanation
The Normal Q-Q plot looks okay for the warpbreaks data.  The histogram appears to be slightly skewed right, but we have a large sample size and a balanced design so it's not too big of a problem.

---
## Conclusion part 1

Write down or email answers to these questions to Kevin

1. When looking at an interaction plot, what would suggest that an interaction is present?

2. Even though we won't do multiple comparisons here, you can still look at the plots to attempt to answer the question of interest: does it appear that any wool-tension combinations resulted in fewer breaks?

3. What are the distribution of the F-statistics for the warpbreaks data?  We found the statistics on slides 18 and 19.  Slide 20 is where I state the distribution of the statistics for the tennis ball data.

4. Take a look back at slide 25.  Is the p-value testing whether method had an effect larger or smaller than the one you obtained in Project 2?  Based on this answer, is there more or less evidence against the null hypothesis that there is no difference in true mean diameter measurements between the methods?  Why does the strength of evidence change? (This last part will require some critical thinking)

5. Do you have any comments or suggestions regrading this activity?


